DSCI 445 Final Report

### 2025-12-15 Paper.Rmd
added eda section

---
title: "DSCI 445 Final Report"
subtitle: "Machine Learning Models on A Bank Marketing Dataset"
author: "Sandra Afrifa, Tigist Kefelew , Mussa Hassen, Waddah Alaahtani"
output: pdf_document
---

@@

*[STREAMLINE Plots and EDA summary HERE]*

>>>>

---
title: "DSCI 445 Final Report"
subtitle: "Machine Learning Models on A Bank Marketing Dataset"
author: "Sandra Afrifa, Tigist Kefelew , Mussa Hassen, Waddah Alqahtani"
output: pdf_document
header-includes:
  - \usepackage{float}
---

@@

## Exploratory Data Analysis (EDA)

### Data Characteristics and Class Imbalance

Our exploratory analysis examined 45,211 client records from the Portuguese banking institution. Analysis of the target variable revealed severe class imbalance, with only 11.7% of clients subscribing to term deposits compared to 88.3% who declined (Figure 1). This imbalance represented a key challenge for predictive modeling.

```{r fig1, echo=FALSE, fig.cap="Class imbalance in term deposit subscriptions. Only 11.7% of clients subscribed, presenting a significant modeling challenge.", fig.align='center', out.width='70%', fig.pos='H'}

ggplot(bank_full, aes(x = y, fill = y)) +
  geom_bar(color = "white", linewidth = 0.4) +
  scale_fill_manual(
    values = c("no" = "grey75", "yes" = "seagreen3")
  ) +
  labs(
    x = "Subscribed?",
    y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank()
  )

### Demographic Patterns

Demographic patterns emerged from the data: retired clients and students were most likely to subscribe, while blue-collar workers were least likely. Higher education levels correlated with increased subscription rates.

```{r fig2, echo=FALSE, fig.cap="Distribution of job categories in the dataset. Administrative, blue-collar, and technician occupations are most common.", fig.align='center', out.width='70%', fig.pos='H'}

ggplot(bank_full, aes(x = job)) +
  geom_bar(fill = "steelblue3") +
  coord_flip() +
  labs(
    x = "Job",
    y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank()
  )

### Historical Patterns

Analysis of previous campaign outcomes revealed powerful patterns: clients with previous successful subscriptions demonstrated approximately 65% likelihood of subscribing again, compared to only 15% among those with previous unsuccessful outcomes.

```{r fig3, echo=FALSE, fig.cap="Previous campaign outcome influence. Clients with previous success show 65% subscription rates compared to 15% for previous failures.", fig.align='center', out.width='70%', fig.pos='H'}

ggplot(bank_full, aes(x = poutcome, fill = y)) +
  geom_bar(position = "fill", color = "white", linewidth = 0.3) +
  scale_fill_manual(
    values = c("no" = "grey75", "yes" = "seagreen3")
  ) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Previous Outcome",
    y = "Proportion Subscribed"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none"
  )

### Campaign Dynamics and Diminishing Returns

Campaign persistence analysis showed a clear downward trend in success rates as contact frequency increased, with substantial decline after five attempts and minimal success beyond thirty contacts.

```{r fig4, echo=FALSE, fig.cap="Diminishing returns from repeated contacts. Success rates drop significantly after 5 contacts.", fig.align='center', out.width='70%', fig.pos='H'}

bank_full %>%
  group_by(campaign) %>%
  summarise(success_rate = mean(y == "yes")) %>%
  ggplot(aes(x = campaign, y = success_rate)) +
  geom_line(color = "steelblue3", linewidth = 1.2) +
  geom_point(color = "darkorange2", size = 2.8) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Number of Contacts",
    y = "Success Rate"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank()
  )
